---
title: "License to Kill (Update: 2014–2024)"
author: "Luis Emilio Muñoz"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)

library(tidyverse)
library(data.table)
library(kableExtra)
library(ggplot2)
library(scales)
library(treemapify)
library(viridis)
library(here)

df <- fread(here::here("individual", "fatal-victims", "clean", "output", "fatal-victims-final.csv"))
```

# 1. Introduction

This report analyzes deaths resulting from police interventions in Puerto Rico between 2014 and 2024. It focuses on **mechanism of death**, **age distribution**, **gender**, **yearly patterns**, and the **proportion of firearm-related deaths** by armed status. 
The analysis is rooted in principles of **epidemiological surveillance**, **data equity**, and **public accountability**.

---

# 2. Distribution by Mechanism of Death
```{r mechanism-frequency}
smmry <- df %>%
  count(mecanismo, name = "count") %>%
  arrange(desc(count))

smmry

treemap_mech <- ggplot(smmry, aes(
  area = count, 
  fill = mecanismo,
  label = paste(mecanismo, "\n", count)
)) +
  geom_treemap() +
  geom_treemap_text(color = "white", place = "center", grow = TRUE) +
  scale_fill_viridis_d(option = "C", end = 0.85) +
  labs(
    title = "Distribution of Deaths by Mechanism",
    caption = "Source: Police intervention deaths dataset (2014–2024)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "none"
  )

treemap_mech
```

---

# 3. Yearly Trends
```{r year-distribution}
year_dist <- df %>%
  count(ano_rip) %>%
  filter(!is.na(ano_rip)) %>%
  arrange(ano_rip)

year_dist_plot <- ggplot(year_dist, aes(x = ano_rip, y = n)) +
  geom_col(fill = viridis::viridis(1, option = "D"), color = "white") +
  geom_text(aes(label = n), vjust = -0.5, size = 5, fontface = "bold") +
  scale_x_continuous(breaks = seq(min(year_dist$ano_rip), max(year_dist$ano_rip), 1)) +
  labs(
    title = "Fatalities per Year (2014–2024)",
    x = "Year",
    y = "Number of Deaths"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

year_dist_plot
```

---

# 4. Firearm Deaths: Armed vs. Unarmed

```{r firearm-proportion}
firearm <- df %>%
    mutate(armed = na_if(armed, "")) %>% #exclude those with unknown information
  filter(mecanismo == "arma de reglamento", !is.na(armed)) %>%
  count(armed) %>%
  mutate(
    Proportion = n / sum(n),
    PercentLabel = paste0(scales::percent(Proportion, 1), "\n", n)
  )

firearm

firearm_pie <- ggplot(firearm, aes(x = "", y = Proportion, fill = armed)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(
    aes(label = PercentLabel),
    position = position_stack(vjust = 0.5),
    color = "white",
    size = 5,
    fontface = "bold"
  ) +
  scale_fill_manual(
    values = c("#F98400", "#666666"),
    name = "Armed Status",
    labels = c("Unarmed", "Armed", "unknown")
  ) +
  labs(
    title = "Firearm Deaths: Armed vs. Unarmed",
    caption = "Among deaths involving firearms"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

firearm_pie
```

---

# 5. Age Distribution (Grouped)

```{r age-distribution}
age <- df %>%
    filter(!is.na(edad))  %>%
  mutate(AgeGroup = cut(
    edad, 
    breaks = c(15, 21, 31, 41, Inf), 
    labels = c("15–20", "21–30", "31–40", "40+"),
    right = FALSE
  ))

age_dist <- age %>%
  count(AgeGroup) %>%
  arrange(AgeGroup)

age_dist

age_plot <- ggplot(age_dist, aes(x = AgeGroup, y = n, fill = AgeGroup)) +
  geom_col(color = "white", show.legend = FALSE) +
  geom_text(aes(label = n), vjust = 1.5, color = "white", size = 6, fontface = "bold") +
  scale_fill_viridis_d(option = "E", end = 0.8) +
  labs(
    title = "Age Distribution of Victims",
    x = "Age Group",
    y = "Number of Deaths"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid = element_blank()
  )

age_plot
```

```{r femincide-prop}
smmry <- df %>%
  count(mecanismo, name = "count") %>%
  arrange(desc(count))

smmry

treemap_mech <- ggplot(smmry, aes(
  area = count, 
  fill = mecanismo,
  label = paste(mecanismo, "\n", count)
)) +
  geom_treemap() +
  geom_treemap_text(color = "white", place = "center", grow = TRUE) +
  scale_fill_viridis_d(option = "C", end = 0.85) +
  labs(
    title = "Distribution of Deaths by Mechanism",
    caption = "Source: Police intervention deaths dataset (2014–2024)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "none"
  )

treemap_mech
```


---
